---
title: "Variation in construction learning across seeds and scales"
author: "Pam Rivière and Sean Trott"
date: "November 3, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(reshape2)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```




# Load Pythia data


Here, we analyze data looking at the average attention each head gives from each token to the previous token.

```{r include=FALSE}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/epistemology/cxn_dev/src/analysis")
directory_path <- "../../data/processed/dais_rs/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list)
nrow(df_pythia_models)

table(df_pythia_models$seed)
max(df_pythia_models$step)
table(df_pythia_models$mpath)


df_pythia_models = df_pythia_models %>%
  mutate(tokens_seen = step * 2100000) %>%
  mutate(model_name = str_remove(mpath, "EleutherAI/"))


```

Merge with judgments:

```{r}
df_judgments = read_csv("../../data/raw/DAIS/generated_pairs_with_results.csv") 

### Merge
df_pythia_models = df_judgments %>%
  mutate(do_sent = DOsentence,
         pd_sent = PDsentence) %>%
  select(-BehavDOpreference) %>%
  inner_join(df_pythia_models) 

nrow(df_pythia_models)
```


# Surprisal over time by seed

Here, we visualize the surprisal curves across seeds/models over time.

```{r raw_surprisal}
### Avg. by step
df_surprisal_by_seed = df_pythia_models %>% 
  group_by(n_params, model_name,tokens_seen, seed_name, revision, step, seed) %>%
  summarise(mean_do_surprisal = mean(do_surprisal_mean),
            mean_po_surprisal = mean(pd_surprisal_mean)) %>%
  mutate(step_modded = step + 1,
         tokens_seen_mod = tokens_seen + 1)

### Avg. across seeds
df_surprisal_avg = df_surprisal_by_seed %>% 
  group_by(n_params,model_name, revision, step, tokens_seen) %>%
  summarise(mean_do_surprisal_across_seeds = mean(mean_do_surprisal),
            mean_po_surprisal_across_seeds = mean(mean_po_surprisal))  %>%
  mutate(step_modded = step + 1,
         tokens_seen_mod = tokens_seen + 1)


ggplot() +
  geom_point(data = df_surprisal_by_seed,
             aes(x = step_modded,
                 y = mean_do_surprisal,
                 color = reorder(model_name, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_surprisal_avg,
            aes(x = step_modded,
                y = mean_do_surprisal_across_seeds,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = df_surprisal_avg,
             aes(x = step_modded,
                 y = mean_do_surprisal_across_seeds,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Mean D.O. Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = 0.15, end = .8))



ggplot() +
  geom_point(data = df_surprisal_by_seed,
             aes(x = step_modded,
                 y = mean_po_surprisal,
                 color = reorder(model_name, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_surprisal_avg,
            aes(x = step_modded,
                y = mean_po_surprisal_across_seeds,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = df_surprisal_avg,
             aes(x = step_modded,
                 y = mean_po_surprisal_across_seeds,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Mean P.D. Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = 0.15, end = .8))



ggplot() +
  geom_point(data = df_surprisal_by_seed,
             aes(x = tokens_seen_mod,
                 y = mean_po_surprisal,
                 color = reorder(model_name, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_surprisal_avg,
            aes(x = tokens_seen_mod,
                y = mean_po_surprisal_across_seeds,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = df_surprisal_avg,
             aes(x = tokens_seen_mod,
                 y = mean_po_surprisal_across_seeds,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Tokens Seen (Log10)",
       y = "Mean P.D. Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 2100000000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = 0.15, end = .8))



```


# Predicting human preferences by num. parameters


```{r}
df_cor_by_seed = df_pythia_models %>% 
  filter(step == 143000) %>%
  group_by(n_params, model_name, seed_name, seed) %>%
  summarise(cor_lr_total = cor(BehavDOpreference, log_odds_total))



### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,model_name) %>%
  summarise(mean_cor_total = mean(cor_lr_total)) 

ggplot(df_cor_by_seed, aes(x = n_params, 
                           y = cor_lr_total,
                           color = reorder(model_name, n_params))) +
  geom_jitter(alpha = 0.7, width = .01) +
  scale_x_log10(labels = scales::comma_format()) +  # Log scale for params
  geom_point(data = df_cor_avg, aes(x = n_params, 
                                    y = mean_cor_total,
                                    color = reorder(model_name, n_params)), 
             size = 4, shape = 17) +
  labs(
    y = "Correlation with human preferences",
    x = "Number of Parameters (log scale)",
    color = ""
  ) +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position="bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = 0.15, end = .8))



```

# Predicting Human Preferences Over Time


```{r}
df_cor_by_seed = df_pythia_models %>% 
  group_by(n_params, model_name, seed_name, revision, step, seed, tokens_seen) %>%
  summarise(cor_lr_total = cor(BehavDOpreference, log_odds_total),
            cor_lr_mean = cor(BehavDOpreference, log_odds_mean)) %>%
  mutate(step_modded = step + 1,
         tokens_seen_mod = tokens_seen + 1)

### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,model_name, revision, step, tokens_seen_mod) %>%
  summarise(mean_cor_total = mean(cor_lr_total),
            mean_cor_mean = mean(cor_lr_mean))  %>%
  mutate(step_modded = step + 1)


ggplot() +
  geom_point(data = df_cor_by_seed,
             aes(x = step_modded,
                 y = cor_lr_total,
                 color = reorder(model_name, n_params)),
             size = 1.2, alpha = 0.3) +
  geom_line(data = df_cor_avg,
            aes(x = step_modded,
                y = mean_cor_total,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = df_cor_avg,
             aes(x = step_modded,
                 y = mean_cor_total,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation with human preference",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8))


ggplot() +
  geom_point(data = df_cor_by_seed,
             aes(x = tokens_seen_mod,
                 y = cor_lr_total,
                 color = reorder(model_name, n_params)),
             size = 1.2, alpha = 0.3) +
  geom_line(data = df_cor_avg,
            aes(x = tokens_seen_mod,
                y = mean_cor_total,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = df_cor_avg,
             aes(x = tokens_seen_mod,
                 y = mean_cor_total,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Tokens Seen (Log10)",
       y = "Correlation with human preference",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 2100000000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8))


```

## Onset, slope, peak

Do larger models show earlier onset of improved correlation with human preferences?

**TODO**: Double-check onset calculation.

```{r onset}
### First, try looking at first onset of a big change, trying out different ratios
### Calculate cross-step ratio
df_diff <- df_cor_by_seed %>%
  group_by(model_name, seed_name) %>%
  arrange(step_modded) %>%
  mutate(
    log_step = log10(step_modded),
    d_ratio = c(NA, diff(cor_lr_total) / diff(log_step))
  ) %>%
  ungroup()

summary(df_diff$d_ratio)

# Define a range of d_ratio thresholds to test
d_ratio_thresholds <- seq(.05, .3, by = 0.01)

# Initialize a list to store the results for each d_ratio threshold
results <- list()

# Iterate over d_ratio thresholds
for (threshold in d_ratio_thresholds) {
  # Filter and find the emergence onset for each threshold
  emergence_onset <- df_diff %>%
    filter(d_ratio > threshold) %>%
    group_by(model_name, seed_name, n_params) %>%  # Group by model size (n_params) and seed_name
    slice_min(step_modded, with_ties = FALSE) %>%
    ungroup()

  # Store the emergence step for each threshold, seed, and model size
  emergence_step_data <- emergence_onset %>%
    select(model_name, seed_name, n_params, step_modded, d_ratio) %>%
    mutate(d_ratio_threshold = threshold)  # Add the threshold as a column

  # Append the result for this threshold to the list
  results[[as.character(threshold)]] <- emergence_step_data
}

# Combine all the results into one data frame
results_df <- bind_rows(results)

results_df %>%
  ggplot(aes(x = d_ratio_threshold,
             y = step_modded,
             color = model_name)) +
  geom_point() +
  theme_minimal()

results_summ = results_df %>%
  group_by(model_name, n_params, seed_name) %>%
  summarise(m_step = mean(step_modded - 1)) 

### Plot average across d_ratio
results_summ %>%
  ggplot(aes(x = n_params,
             y = m_step,
             color = reorder(model_name, n_params))) +
  geom_jitter(size = 5, alpha = .7, width = .05) +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8)) +
  scale_x_log10() +
  scale_y_log10() +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Number of Parameters (Log10)",
       y = "Max. Deriv. Step (Log10)",
       color = "") 


### For each order magnitude in model size, how much sooner do we expect first onset on avearge?
mod_ratio = lmer(data = results_summ,
                 log10(m_step) ~ log10(n_params) + (1|seed_name))
summary(mod_ratio)



### Now fit with GAMs
library(mgcv)

# Fit GAM across all models/seeds
df_cor_by_seed$log_step = log10(df_cor_by_seed$step_modded)
gam_all <- gam(cor_lr_total ~ s(log_step), data = df_cor_by_seed)

# Summary of the GAM
summary(gam_all)

# Plot the GAM curve
plot(gam_all)


### Now plot it over the original models
newdat_all <- tibble(
  log_step = seq(min(df_cor_by_seed$log_step),
                 max(df_cor_by_seed$log_step),
                 length.out = 200)
) %>%
  mutate(step_modded = 10^log_step,
         gam_pred = predict(gam_all, newdata = .))


# Plot seed-level + average + OVERALL GAM across all facets
summary_avg = df_cor_by_seed %>%
  group_by(model_name, n_params, step_modded) %>%
  summarise(
    mean_across_seeds = mean(cor_lr_total)
  )


df_cor_by_seed %>%
  ggplot(aes(x = step_modded,
             y = cor_lr_total,
             color = factor(seed_name))) +
  geom_line(size = .6, alpha = .4) +   # individual curves
  geom_line(data = summary_avg, 
            aes(x = step_modded, y = mean_across_seeds), 
            color = "black", size = 2) +      # per-model average
  geom_line(data = newdat_all, 
            aes(x = step_modded, y = gam_pred), 
            inherit.aes = FALSE, 
            alpha = .6,
            color = "red", size = 1.5) +        # overall GAM overlay
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation with Human Preferences",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000, linetype = "dotted", size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(9, option = "mako", 
                                                   begin = 0.8, end = 0.15)) + 
  facet_wrap(~reorder(model_name, n_params))






```


Do larger models show a higher peak performance?

```{r peak}
df_peak <- df_cor_by_seed %>%
  group_by(model_name, seed_name, n_params) %>%
  summarise(max_cor = max(cor_lr_total))

df_peak %>%
  group_by(n_params, model_name) %>%
  summarise(mean_max_cor = mean(max_cor))

summary(lmer(data = df_peak, 
        max_cor ~ log10(n_params) + (1|seed_name)))


df_peak %>%
  ggplot(aes(x = n_params,
             y = max_cor,
             color = reorder(model_name, n_params))) +
  geom_jitter(size = 5, alpha = .7, width = .05) +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8)) +
  scale_x_log10() +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Number of Parameters (Log10)",
       y = "Max. Peak",
       color = "") 

```

What about reduced slope?

```{r slope}
df_cor_by_seed <- df_cor_by_seed %>%
  mutate(log_step = log10(step_modded))

# Fit linear models by model + seed
slopes_df <- df_cor_by_seed %>%
  group_by(model_name, n_params, seed_name) %>%
  nest() %>%
  mutate(
    fit = map(data, ~ lm(cor_lr_total ~ log_step, data = .x)),
    tidied = map(fit, broom::tidy)
  ) %>%
  unnest(tidied) %>%
  filter(term == "log_step") %>%
  select(model_name, seed_name, estimate, std.error, statistic, p.value)


slopes_df %>%
  group_by(n_params, model_name) %>%
  summarise(mean_slope = mean(estimate))


summary(lmer(data = slopes_df, 
        estimate ~ log10(n_params) + (1|seed_name)))

slopes_df %>%
  ggplot(aes(x = n_params,
             y = estimate,
             color = reorder(model_name, n_params))) +
  geom_jitter(size = 5, alpha = .7, width = .05) +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8)) +
  scale_x_log10() +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Number of Parameters (Log10)",
       y = "Slope (Correlation ~ Step)",
       color = "") 

```

## Wisdom of the seeds?

Here, we ask whether averaging across each seed's estimates for a given sentence pair at a certain checkpoint improves performance. We find that the "ensemble" estimate is indeed more correlated with human judgments, on average.

```{r}
### Wisdom of the crowd
df_wc = df_pythia_models %>%
  group_by(model_name, step, do_sent, pd_sent, BehavDOpreference) %>%
  summarise(mean_lr = mean(log_odds_total)) %>%
  ungroup() %>%
  group_by(model_name, step) %>%
  mutate(cor_wc = cor(mean_lr, BehavDOpreference))


### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,model_name, revision, step, tokens_seen_mod) %>%
  summarise(mean_cor_total = mean(cor_lr_total),
            max_cor_total = max(cor_lr_total))  %>%
  mutate(step_modded = step + 1)


df_joined = df_wc %>%
  inner_join(df_cor_avg)

summary(df_joined$max_cor_total)
summary(df_joined$cor_wc)

df_joined %>%
  mutate(wisdom_benefit_to_max = cor_wc - max_cor_total,
         wisdom_benefit_to_mean = cor_wc - mean_cor_total) %>%
  filter(step == 143000) %>%
  select(model_name, wisdom_benefit_to_max, wisdom_benefit_to_mean,
         cor_wc, max_cor_total, mean_cor_total)


df_joined %>%
  pivot_longer(cols = c(mean_cor_total, cor_wc, max_cor_total),
               names_to = "metric",
               values_to = "correlation") %>%
  ggplot(aes(x = tokens_seen_mod, y = correlation, color = metric)) +
  geom_line(linewidth = 1) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_color_manual(values = viridisLite::viridis(3, option = "mako", 
                                                    begin = .15, end = .8),
                     labels = c("mean_cor_total" = "Mean Across Seeds", 
               "cor_wc" = "Ensemble",
               "max_cor_total" = "Max Across Seeds")) +
  theme_minimal() +
  labs(x = "Tokens Seen (log scale)",
       y = "Correlation with Human Preferences",
       color = "") +
  facet_wrap(~reorder(model_name, n_params)) +
  theme(text = element_text(size = 15),
        legend.position = "bottom")



####### Comparing to individual seeds
df_summary = df_comparison %>%
  group_by(model_name, n_params, step_modded, tokens_seen_mod) %>%
  summarise(
    ensemble_cor = first(cor_wc),
    mean_individual = mean(cor_lr_total),
    max_individual = max(cor_lr_total),
    min_individual = min(cor_lr_total),
    exceeds_mean = ensemble_cor > mean_individual,
    n_seeds_beaten = sum(cor_wc > cor_lr_total),
    total_seeds = n()
  )

# Visualization 1: Ensemble vs. Mean/Max over time
df_summary %>%
  ggplot(aes(x = tokens_seen_mod)) +
  geom_ribbon(aes(ymin = min_individual, ymax = max_individual), 
              alpha = 0.2, fill = "grey50") +
  geom_line(aes(y = mean_individual, color = "Mean Individual"), 
            linewidth = 1, linetype = "dashed") +
  geom_line(aes(y = ensemble_cor, color = "Ensemble"), 
            linewidth = 1.2) +
  geom_point(aes(y = ensemble_cor, color = "Ensemble"), size = 2) +
  scale_x_log10() +
  scale_color_manual(
    values = c("Mean Individual" = "#0072B2", "Ensemble" = "#D55E00")
  ) +
  facet_wrap(~reorder(model_name, n_params), scales = "free_x") +
  theme_minimal() +
  labs(x = "Tokens Seen (log scale)",
       y = "Correlation with Human Preferences",
       color = NULL,
       subtitle = "Grey ribbon shows range of individual seeds")


### How many seeds for ensemble beat?
df_summary %>%
  ggplot(aes(x = tokens_seen_mod, y = n_seeds_beaten / total_seeds)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_x_log10() +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~reorder(model_name, n_params), scales = "free_x") +
  theme_minimal() +
  labs(x = "Tokens Seen (log scale)",
       y = "% Seeds Beaten by Ensemble",
       subtitle = "Red line at 50% (better than median)")


#### Sig. testing
df_ttest = df_comparison %>%
  group_by(model_name, n_params, step_modded, tokens_seen_mod) %>%
  summarise(
    mean_benefit = mean(wisdom_benefit_to_seed),
    se_benefit = sd(wisdom_benefit_to_seed) / sqrt(n()),
    ci_lower = mean_benefit - 1.96 * se_benefit,
    ci_upper = mean_benefit + 1.96 * se_benefit,
    p_value = t.test(wisdom_benefit_to_seed, mu = 0, alternative = "greater")$p.value,
    is_sig = p_value < 0.05
  )

# Visualization 1: Benefit over time with confidence intervals
df_ttest %>%
  ggplot(aes(x = tokens_seen_mod, y = mean_benefit)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.3, fill = "#D55E00") +
  geom_line(linewidth = 1, color = "#D55E00") +
  geom_point(aes(shape = is_sig), size = 3, color = "#D55E00") +
  scale_x_log10() +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     labels = c("TRUE" = "p < .05", "FALSE" = "n.s.")) +
  facet_wrap(~reorder(model_name, n_params), scales = "free_x") +
  theme_minimal() +
  labs(x = "Tokens Seen (log scale)",
       y = "Ensemble Benefit (Δ Correlation)",
       shape = "Significance",
       subtitle = "Ribbon shows 95% CI")


df_comparison = df_cor_by_seed %>%
  ungroup() %>%
  select(model_name, step_modded, seed_name, cor_lr_total) %>%
  inner_join(df_joined) %>%
  mutate(wisdom_benefit_to_seed = cor_wc - cor_lr_total)

df_comparison %>%
  group_by(model_name, step_modded) %>%
  summarise(
    mean_benefit = mean(wisdom_benefit_to_seed),
    sd_benefit = sd(wisdom_benefit_to_seed),
    t_stat = t.test(wisdom_benefit_to_seed, mu = 0, alternative = "greater")$statistic,
    p_value = t.test(wisdom_benefit_to_seed, mu = 0, alternative = "greater")$p.value,
    n_seeds = n()
  )

model <- lmer(wisdom_benefit_to_seed ~ log10(step_modded) + log10(n_params) + 
              (1|seed_name), 
              data = df_comparison)
summary(model)

  


### Plot final-step
df_pythia_models %>%
  filter(step == 143000) %>%
  group_by(model_name, n_params, do_sent, pd_sent, 
           classification, BehavDOpreference) %>%
  summarise(mean_lr = mean(log_odds_total)) %>%
  ggplot(aes(x = BehavDOpreference,
             y = mean_lr,
             color = classification)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm") +
  labs(x = "DO Preference",
       y = "Log Ratio (Ensemble)",
       fill = "") +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  theme_ridges() +
  facet_wrap(~reorder(model_name, n_params))
```





# Effect of explanatory variables




## Sensitivity to classification


**TODO**: Does the "peak" at ~512 suggest that this is when models actually learn the DO construction in general (i.e., around ~256 they're starting to capture human preferences, but at ~512 they've learned DOs, and then after that they learn which verbs go with which).

```{r classification}
### final step 
df_pythia_models %>%
  filter(step == 143000) %>%
  # filter(mpath == "EleutherAI/pythia-70m") %>%
  ggplot(aes(x = log_odds_total,
             y = seed_name,
             fill = recipient_id)) +
  geom_density_ridges(alpha = .6, color = NA) +
  labs(x = "Log Ratio",
       y = "Seed",
       fill = "") +
  scale_fill_manual(values = viridisLite::viridis(5, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  theme_ridges() +
  facet_wrap(~reorder(mpath, n_params))


## Over pretraining
coef_results = df_pythia_models %>%
  group_by(model_name, n_params, seed_name, step) %>%
  group_modify(~broom::tidy(lm(log_odds_total ~ classification, data = .x))) %>%
  filter(term != "(Intercept)")

### Avg. across seeds
coef_avg = coef_results %>% 
  group_by(n_params,model_name, step) %>%
  summarise(mean_estimate = mean(estimate))  %>%
  mutate(step_modded = step + 1)


coef_results %>%
  mutate(step_modded = step + 1) %>%
  ggplot(aes(x = step_modded, 
             y = estimate, color = reorder(model_name, n_params))) +
  geom_point(alpha = .5) +
  geom_line(data = coef_avg,
            aes(x = step_modded,
                y = mean_estimate,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = coef_avg,
             aes(x = step_modded,
                 y = mean_estimate,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Coefficient (Classification: Non-alternating)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8))




coef_results = df_pythia_models %>%
  group_by(model_name, n_params, seed_name, step) %>%
  group_modify(~broom::tidy(lm(log_odds_total ~ recipient_id, data = .x))) %>%
  filter(term != "(Intercept)")

### Avg. across seeds
coef_avg = coef_results %>% 
  group_by(n_params,model_name, step) %>%
  summarise(mean_estimate = mean(estimate))  %>%
  mutate(step_modded = step + 1)


coef_results %>%
  mutate(step_modded = step + 1) %>%
  ggplot(aes(x = step_modded, 
             y = estimate, color = reorder(model_name, n_params))) +
  geom_point(alpha = .5) +
  geom_line(data = coef_avg,
            aes(x = step_modded,
                y = mean_estimate,
                color = reorder(model_name, n_params)),
            size = 1.3) +
  geom_point(data = coef_avg,
             aes(x = step_modded,
                 y = mean_estimate,
                 color = reorder(model_name, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Coefficient (Classification: Non-alternating)",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = .8)) +
  facet_wrap(~term)

```


## Sensitivity to theme

```{r}
df_judgments %>%
  group_by(theme_type) %>%
  summarise(mean_do = mean(BehavDOpreference))
  ungroup() %>%
  mutate(classification = fct_relevel(classification, 'non-alternating')) %>%
  ggplot(aes(x = recipient_id,  y = empirical_stat, fill = theme_type)) +
    geom_bar(stat = 'identity', position=position_dodge(width = 0.9)) +
    geom_errorbar(aes(ymin = ci_lower, ymax=ci_upper), width = 0, position=position_dodge(width = 0.9)) +
    labs(x = 'recipient type', y = 'average DO preference') +
    guides(x = guide_axis(angle = 90)) +
    scale_fill_colorblind()+
    facet_wrap(~classification) +
    theme_few() +
    theme(aspect.ratio = 1, legend.position = 'top')

```


## Sensitivity to recipient information


**TODO**: Figure out how to handle pronouns in the linear model. (See: https://github.com/taka-yamakoshi/neural_constructions/blob/master/DAIS/analysis/behavioral.Rmd)

**TODO**...


```{r}
df_judgments = read_csv("../../data/raw/DAIS/generated_pairs_with_results.csv") %>%
  mutate(
    # Recode pronoun as 'extrashort_definite' before separating
    recipient_id = case_when(
      recipient_id == 'shortDefinite' ~ 'short_definite',
      recipient_id == 'longDefinite' ~ 'long_definite',
      recipient_id == 'shortIndefinite' ~ 'short_indefinite',
      recipient_id == 'longIndefinite' ~ 'long_indefinite',
      recipient_id == 'pronoun' ~ 'extrashort_definite'
    )
  ) %>%
  separate(recipient_id, 
           into = c('recipient_length', 'recipient_definiteness'), 
           sep = '_') %>%
  mutate(
    # Set reference levels (optional - R will use alphabetical by default)
    recipient_length = factor(recipient_length, 
                             levels = c("extrashort", "short", "long")),
    recipient_definiteness = factor(recipient_definiteness,
                                   levels = c("definite", "indefinite"))
  )

df_merged = df_judgments %>%
  mutate(do_sent = DOsentence,
         pd_sent = PDsentence) %>%
  inner_join(df_pythia_models) 

coef_results = df_merged %>%
  group_by(mpath, n_params, seed_name) %>%
  group_modify(~broom::tidy(lm(log_odds_total ~ classification + recipient_length +
                                 recipient_definiteness, data = .x))) %>%
  filter(term != "(Intercept)")


coef_summary <- coef_results %>%
  group_by(mpath, n_params, term) %>%
  summarise(
    mean_est = mean(estimate),
    sd_est = sd(estimate),
    se_est = sd(estimate) / sqrt(n())
  )

library(ggstance)

ggplot(coef_results, aes(y = term, x = estimate)) +
  geom_pointrange(data = coef_summary,
                  aes(x = mean_est, 
                      xmin = mean_est - 1.96*se_est, 
                      xmax = mean_est + 1.96*se_est),
                  size = 0.8, linewidth = 1.2) +
  geom_point(aes(color = seed_name), 
             position = position_jitter(height = 0.2),
             size = 2, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Coefficient estimate",
    y = "Model / Term",
    color = "Seed"
  ) +
  theme_minimal() +
  facet_wrap(~mpath)



ggplot(coef_results, aes(x = seed_name, y = estimate, group = term, color = term)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~mpath, scales = "free_x") +
  labs(
    x = "Seed",
    y = "Coefficient estimate",
    color = "Term"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




mod = lmer(data = df_merged %>%
             filter(recipient_id %in% c("longIndefinite", "shortIndefinite")),
           log_odds_mean ~ recipient_id + (1 + classification|seed_name))
summary(mod)


mod = lmer(data = df_merged %>%
             filter(recipient_id %in% c("longDefinite", "shortDefinite")),
           log_odds_mean ~ recipient_id + (1 + classification|seed_name))
summary(mod)


df_merged %>%
  group_by(mpath, seed_name, recipient_length) %>%
  summarise(mean_lr_total = mean(log_odds_total),
            se_lr_total = sd(log_odds_total)/sqrt(n())) %>%
  ggplot(aes(x = reorder(recipient_length, mean_lr_total),
             y = mean_lr_total)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(x = reorder(recipient_length, mean_lr_total),
                    ymin = mean_lr_total - se_lr_total,
                    ymax = mean_lr_total + se_lr_total),
                width = .2) +
  theme_minimal() +
  facet_grid(~seed_name~mpath)

# Calculate mean across seeds for each model/length/definiteness combo
df_seed_means <- df_merged %>%
  group_by(mpath, n_params, recipient_length, recipient_definiteness) %>%
  summarise(
    mean_lr_total = mean(log_odds_total),
    se_lr_total = sd(log_odds_total) / sqrt(n()),
    .groups = "drop"
  )

# Calculate individual seed values
df_seed_values <- df_merged %>%
  group_by(mpath, n_params, seed_name, recipient_length, recipient_definiteness) %>%
  summarise(mean_lr_total = mean(log_odds_total), .groups = "drop")


ggplot() +
  # Individual seeds - smaller points with jitter
  geom_point(data = df_seed_values,
             aes(x = mean_lr_total, 
                 y = reorder(recipient_length, mean_lr_total),
                 color = recipient_definiteness),
             position = position_jitter(height = 0.15, seed = 42),
             size = 2, alpha = 0.5) +
  # Mean across seeds - larger points with error bars
  geom_pointrange(data = df_seed_means,
                  aes(x = mean_lr_total,
                      xmin = mean_lr_total - se_lr_total,
                      xmax = mean_lr_total + se_lr_total,
                      y = reorder(recipient_length, mean_lr_total),
                      color = recipient_definiteness),
                  size = 1, linewidth = 1.2,
                  position = position_dodge(width = 0.3)) +  # Changed to width
  labs(
    x = "Log odds (DO preference)",
    y = "Recipient length",
    color = "Recipient\nDefiniteness"
  ) +
  facet_wrap(~mpath, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "right")



#### Human baseline

mod = lmer(data = df_judgments,
           BehavDOpreference ~ classification + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(recipient_id %in% c("longIndefinite", "shortIndefinite")),
           BehavDOpreference ~ recipient_id + (1 + recipient_id|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(recipient_id %in% c("longDefinite", "shortDefinite")),
           BehavDOpreference ~ recipient_id + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(theme_type != "something"),
           BehavDOpreference ~ theme_type + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments,
           BehavDOpreference ~ classification + (1 + classification|verb_id))
summary(mod)


df_pythia_models %>%
  mutate(lr = pd_surprisal_total - do_surprisal_total) %>%
  ggplot(aes(x = lr,
             log_odds_total)) +
  geom_point()
```




```{r}
df_cor_by_seed = df_merged %>% 
  group_by(n_params, mpath, seed_name, revision, step, seed) %>%
  summarise(cor_lr_total = cor(BehavDOpreference, log_odds_total),
            cor_lr_gpt2 = cor(`gpt2-large_ratio`, log_odds_total),
            cor_lr_ngram = cor(ngram_ratio, log_odds_total)) %>%
  mutate(step_modded = step + 1)

### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,mpath, revision, step) %>%
  summarise(mean_cor_total = mean(cor_lr_total),
            mean_cor_gpt2 = mean(cor_lr_gpt2),
            mean_cor_ngram = mean(cor_lr_ngram))  %>%
  mutate(step_modded = step + 1)


ggplot() +
  geom_point(data = df_cor_by_seed,
             aes(x = step_modded,
                 y = cor_lr_gpt2,
                 color = reorder(mpath, n_params)),
             size = 1.2, alpha = 0.3) +
  geom_line(data = df_cor_avg,
            aes(x = step_modded,
                y = mean_cor_gpt2,
                color = reorder(mpath, n_params)),
            size = 1.3) +
  geom_point(data = df_cor_avg,
             aes(x = step_modded,
                 y = mean_cor_gpt2,
                 color = reorder(mpath, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation with GPT-2 (large) preference",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = 1))


ggplot() +
  geom_point(data = df_cor_by_seed,
             aes(x = step_modded,
                 y = cor_lr_ngram,
                 color = reorder(mpath, n_params)),
             size = 1.2, alpha = 0.3) +
  geom_line(data = df_cor_avg,
            aes(x = step_modded,
                y = mean_cor_ngram,
                color = reorder(mpath, n_params)),
            size = 1.3) +
  geom_point(data = df_cor_avg,
             aes(x = step_modded,
                 y = mean_cor_ngram,
                 color = reorder(mpath, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation with n-gram preference",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = 1))


cor.test(df_judgments$`gpt2-large_ratio`,
         df_judgments$BehavDOpreference)

cor.test(df_judgments$ngram_ratio,
         df_judgments$BehavDOpreference)
```




# Convergence across models

**TODO**: Two questions here:

- First, how does sentence-level LR correlate at each time step between model pairs, and how does that change? 
- Second, how does the correlation with human ground truth correlate between model pairs? (Analogize to developmental trajectory of 1-back attention.)


```{r mds}
### Sentence-level vs. trajectory over time?
df_wide <- df_pythia_models %>%
  filter(mpath != "EleutherAI/pythia-160m") %>%
  ungroup() %>%
  mutate(model_id = paste(mpath, "-", seed_name)) %>%
  select(step, model_id, do_sent, pd_sent, log_odds_total) %>%
  pivot_wider(
    names_from = model_id,
    values_from = log_odds_total,
    id_cols = c(step, do_sent, pd_sent)  # explicitly specify id columns
  ) 

correlations_long <- df_wide %>%
  group_by(step) %>%
  summarise(
    cor_matrix = list(
      cor(across(starts_with("EleutherAI")), use = "pairwise.complete.obs") %>%
        as.data.frame() %>%
        rownames_to_column("model1") %>%
        pivot_longer(-model1, names_to = "model2", values_to = "correlation")
    )
  ) %>%
  unnest(cor_matrix) %>%
  mutate(same_model = model1 == model2) %>%
  filter(same_model == FALSE) # %>%
  # separate(model1, into = c("model1", "seed1"), sep = " - ") %>%
  # separate(model2, into = c("model2", "seed2"), sep = " - ") %>%
  # mutate(same_architecture = model1 == model2)

correlations_long %>%
  filter(step %in% c(0, 1, 4, 8, 16, 64, 256, 143000)) %>%
  ggplot(aes(x = model1, y = model2, fill = correlation)) +
  geom_tile() +
  coord_fixed() +
  theme_minimal(base_size = 10) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limit = c(-1, 1),
    name = "Corr"
  ) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(size = 10),
    legend.position = "right"
  ) +
  coord_fixed() +
  facet_wrap(~step, ncol = 4)



df_params = df_pythia_models %>%
  mutate(model = mpath, seed = seed_name) %>%
  dplyr::select(model, seed, n_params) %>%
  distinct() 

##### Multi-dimensional scaling
mds_by_step <- df_wide %>%
  # drop_na() %>%
  group_by(step) %>%
  summarise(
    mds_df = list({
      cor_mat <- cor(dplyr::select(cur_data(), starts_with("EleutherAI")), use = "pairwise.complete.obs")
      dist_mat <- as.dist(1 - cor_mat)
      mds <- cmdscale(dist_mat, k = 2)
      as_tibble(mds, .name_repair = "unique") %>%
        mutate(model_id = colnames(cor_mat)) 
    }),
    .groups = "drop"
  ) %>%
  unnest(mds_df) %>%
  separate(model_id, into = c("model", "seed"), sep = " - ") %>%
  rename(x = `...1`, y = `...2`) %>%
  inner_join(df_params)



mds_by_step %>%
  filter(step %in% c(0, 1, 4, 8, 16, 64, 256, 143000)) %>%
  ggplot(aes(x, y, 
           color = reorder(model, n_params))) +
  geom_point() +
  facet_wrap(~ step, ncol = 4) +
  theme_bw() +
  labs(x = "MDS 1",
       y = "MDS 2",
       color = "") +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                   begin = 0.8, end = 0.15)) 



df_wide <- df_pythia_models %>%
  mutate(step_modded = step + 1) %>%
  ungroup() %>%
  mutate(model_id = paste(mpath, "-", seed_name)) %>%
  dplyr::select(step_modded, model_id, seed_name, do_sent, pd_sent, log_odds_total) %>%
  # group_by(model_id, step_modded) %>%
  # summarise(max_attn = mean(max_attn, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = model_id,
    values_from = log_odds_total
  )

model_id_to_params <- df_pythia_models %>%
  dplyr::select(mpath, seed_name, n_params) %>%
  distinct() %>%
  mutate(model_id = paste(mpath, "-", seed_name)) %>%
  distinct(model_id, n_params) %>%
  arrange(n_params) %>%
  pull(model_id)


### 
cor_long <- df_wide %>%
  # drop the step column; keep only model time series
  dplyr::select(-step_modded, -do_sent, -pd_sent) %>%
  cor(use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("Var1") %>%
  pivot_longer(
    cols = -Var1,
    names_to = "Var2",
    values_to = "value"
  ) %>%
  mutate(
    Var1 = factor(Var1, levels = model_id_to_params),
    Var2 = factor(Var2, levels = model_id_to_params)
  )



### Make correlation matrix
cor_long %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  coord_fixed() +
  theme_minimal(base_size = 10) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limit = c(-1, 1),
    name = "Corr"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        axis.title = element_blank()) +
  coord_fixed()


### Avg. correlation within models
# extract model name without seed
cor_long <- cor_long %>%
  # extract everything before the last " - " as the model
  mutate(
    model1 = str_replace(Var1, " - .*", ""),
    model2 = str_replace(Var2, " - .*", ""),
    same_model = model1 == model2
  ) %>% 
  mutate(same_exact = Var1 == Var2) %>%
  filter(same_exact == FALSE)

cor_long = cor_long %>%
  left_join(df_pythia_models %>% select(model, n_params) %>% distinct(),
            by = c("model1" = "model")) %>%
  rename(n_params1 = n_params) %>%
  left_join(df_pythia_models %>% select(model, n_params) %>% distinct(),
            by = c("model2" = "model")) %>%
  rename(n_params2 = n_params) 


cor_long$cor = cor_long$value

summary(lm(data = cor_long,
             cor ~ same_model + log10(n_params1) + log10(n_params2)))


summary(lm(data = cor_long %>% filter(same_model == FALSE),
             cor ~ log10(n_params1) + log10(n_params2)))




cor_long %>%
  group_by(same_model) %>%
  summarise(mean_cor = mean(cor),
            sd_cor = sd(cor))

cor_long %>%
  group_by(same_model, n_params1) %>%
  summarise(mean_cor = mean(cor))

cor_long %>%
  mutate(abs_diff_params = round(abs(log10(n_params2 - log10(n_params1))), 2)) %>%
  mutate(diff_params = abs(n_params2 - n_params1)) %>%
  filter(same_model == FALSE) %>%
  group_by(diff_params) %>%
  summarise(mean_cor = mean(cor))






# Calculate centroids for each model size
centroids <- mds_by_step %>%
  group_by(step, n_params) %>%
  summarise(
    centroid_x = mean(x),
    centroid_y = mean(y),
    .groups = "drop"
  )

# Calculate separation ratio for each model size
separation_stats <- centroids %>%
  rowwise() %>%
  mutate(
    # Distance to other centroids
    inter_dist = {
      current_params <- n_params
      other_centroids <- centroids %>% filter(n_params != current_params)
      mean(sqrt((other_centroids$centroid_x - centroid_x)^2 + 
                (other_centroids$centroid_y - centroid_y)^2))
    },
    # Within-cluster spread
    within_dist = {
      current_params <- n_params
      current_x <- centroid_x
      current_y <- centroid_y
      mds_by_step %>% 
        filter(n_params == current_params) %>%
        summarise(spread = mean(sqrt((x - current_x)^2 + (y - current_y)^2))) %>%
        pull(spread)
    }
  ) %>%
  mutate(sep_ratio = inter_dist / within_dist) %>%
  ungroup()

print(separation_stats)
```

