---
title: "Variation in construction learning across seeds and scales"
author: "Pam Rivière and Sean Trott"
date: "November 3, 2025"
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi = 300, fig.format = "pdf")
```


```{r include=FALSE}
library(tidyverse)
library(lme4)
library(viridis)
library(ggridges)
library(lmerTest)
library(ggrepel)
library(ggcorrplot)
library(reshape2)

all_colors <- viridis::viridis(10, option = "mako")
my_colors <- all_colors[c(3, 5, 7)]  # Selecting specific colors from the palette
```




# Load Pythia data


Here, we analyze data looking at the average attention each head gives from each token to the previous token.

```{r include=FALSE}
# setwd("/Users/seantrott/Dropbox/UCSD/Research/NLMs/development/cxn_dev/src/analysis")
directory_path <- "../../data/processed/dais_rs/"
csv_files <- list.files(path = directory_path, pattern = "*.csv", full.names = TRUE)
csv_list <- csv_files %>%
  map(~ read_csv(.))
df_pythia_models <- bind_rows(csv_list)
nrow(df_pythia_models)

table(df_pythia_models$seed)
max(df_pythia_models$step)
table(df_pythia_models$mpath)

```


# Surprisal over time by seed

Here, we visualize the surprisal curves across seeds/models over time.

```{r}
### Avg. by step
df_surprisal_by_seed = df_pythia_models %>% 
  group_by(n_params, mpath, seed_name, revision, step, seed) %>%
  summarise(mean_do_surprisal = mean(do_surprisal_mean),
            mean_po_surprisal = mean(pd_surprisal_mean)) %>%
  mutate(step_modded = step + 1)

### Avg. across seeds
df_surprisal_avg = df_surprisal_by_seed %>% 
  group_by(n_params,mpath, revision, step) %>%
  summarise(mean_do_surprisal_across_seeds = mean(mean_do_surprisal),
            mean_po_surprisal_across_seeds = mean(mean_po_surprisal))  %>%
  mutate(step_modded = step + 1)


ggplot() +
  geom_point(data = df_surprisal_by_seed,
             aes(x = step_modded,
                 y = mean_do_surprisal,
                 color = reorder(mpath, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_surprisal_avg,
            aes(x = step_modded,
                y = mean_do_surprisal_across_seeds,
                color = reorder(mpath, n_params)),
            size = 1.3) +
  geom_point(data = df_surprisal_avg,
             aes(x = step_modded,
                 y = mean_do_surprisal_across_seeds,
                 color = reorder(mpath, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Mean D.O. Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                    begin = 0.8, end = 0.15))



ggplot() +
  geom_point(data = df_surprisal_by_seed,
             aes(x = step_modded,
                 y = mean_po_surprisal,
                 color = reorder(mpath, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_surprisal_avg,
            aes(x = step_modded,
                y = mean_po_surprisal_across_seeds,
                color = reorder(mpath, n_params)),
            size = 1.3) +
  geom_point(data = df_surprisal_avg,
             aes(x = step_modded,
                 y = mean_po_surprisal_across_seeds,
                 color = reorder(mpath, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Mean P.D. Surprisal across seeds",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "mako", 
                                                    begin = 0.8, end = 0.15))




df_pythia_models %>%
  ggplot(aes(x = log_odds_total,
             y = log_odds_mean)) +
  geom_point(alpha = .5) +
  theme_minimal() +
  facet_wrap(~seed_name)

```


# Predicting human preferences by num. parameters


```{r}
df_cor_by_seed = df_pythia_models %>% 
  filter(step == 143000) %>%
  group_by(n_params, mpath, seed_name, seed) %>%
  summarise(cor_lr_total = cor(BehavDOpreference, log_odds_total))



### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,mpath) %>%
  summarise(mean_cor_total = mean(cor_lr_total)) 

ggplot(df_cor_by_seed, aes(x = n_params, y = cor_lr_total)) +
  geom_point(alpha = 0.7) +
  scale_x_log10(labels = scales::comma_format()) +  # Log scale for params
  geom_line(data = df_cor_avg, aes(x = n_params, y = mean_cor_total), 
            color = "red", size = 1.2) +
  geom_point(data = df_cor_avg, aes(x = n_params, y = mean_cor_total), 
             color = "red", size = 4, shape = 17) +
  labs(
    y = "Correlation with human preferences",
    x = "Number of Parameters (log scale)"
  ) +
  theme_minimal() +
  theme(text = element_text(size = 15),
        legend.position="bottom") 


```

# Predicting Human Preferences Over Time

**TODO**: Does "Total" LR difference just ubild in length effect?

```{r}
df_cor_by_seed = df_pythia_models %>% 
  group_by(n_params, mpath, seed_name, revision, step, seed) %>%
  summarise(cor_lr_total = cor(BehavDOpreference, log_odds_total),
            cor_lr_mean = cor(BehavDOpreference, log_odds_mean)) %>%
  mutate(step_modded = step + 1)

### Avg. across seeds
df_cor_avg = df_cor_by_seed %>% 
  group_by(n_params,mpath, revision, step) %>%
  summarise(mean_cor_total = mean(cor_lr_total),
            mean_cor_mean = mean(cor_lr_mean))  %>%
  mutate(step_modded = step + 1)


ggplot() +
  geom_point(data = df_cor_by_seed,
             aes(x = step_modded,
                 y = cor_lr_total,
                 color = reorder(mpath, n_params)),
             size = 2, alpha = 0.3) +
  geom_line(data = df_cor_avg,
            aes(x = step_modded,
                y = mean_cor_total,
                color = reorder(mpath, n_params)),
            size = 1.3) +
  geom_point(data = df_cor_avg,
             aes(x = step_modded,
                 y = mean_cor_total,
                 color = reorder(mpath, n_params)),
             size = 4, alpha = 0.4) +
  theme_minimal() +
  labs(x = "Training Step (Log10)",
       y = "Correlation with human preference",
       color = "") +
  scale_x_log10() +
  geom_vline(xintercept = 1000,
             linetype = "dotted",
             size = 1.2) +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  scale_color_manual(values = viridisLite::viridis(4, option = "magma", 
                                                    begin = .15, end = 1))
```


# Final step behavhior

```{r}
df_pythia_models %>%
  filter(step == 143000) %>%
  # filter(mpath == "EleutherAI/pythia-70m") %>%
  ggplot(aes(x = log_odds_total,
             y = seed_name,
             fill = classification)) +
  geom_density_ridges(alpha = .6, color = NA) +
  labs(x = "Log Ratio",
       y = "Seed") +
  scale_fill_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  theme_ridges() +
  facet_wrap(~mpath)

df_pythia_models %>%
  filter(step == 143000) %>%
  # filter(mpath == "EleutherAI/pythia-70m") %>%
  ggplot(aes(x = BehavDOpreference,
             y = log_odds_total,
             color = classification)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm") +
  facet_wrap(~mpath + seed_name) +
  theme_minimal() +
  scale_color_manual(values = viridisLite::viridis(2, option = "mako", 
                                                    begin = 0.8, end = 0.15)) +
  labs(y = "Surprisal Ratio (PO - DO)",
       x = "DO Preference")





mod = lmer(data = df_pythia_models %>%
             filter(step == 143000) %>% 
             filter(mpath == "EleutherAI/pythia-70m"),
           log_odds_total ~ classification + (1 + classification|seed_name))
summary(mod)


mod = lmer(data = df_pythia_models,
           BehavDOpreference ~ classification + log_odds_total + (1 + classification|seed_name))
summary(mod)


```


# Effect of other variables


**TODO**: Figure out how to handle pronouns in the linear model. (See: https://github.com/taka-yamakoshi/neural_constructions/blob/master/DAIS/analysis/behavioral.Rmd)

```{r}
df_judgments = read_csv("../../data/raw/DAIS/generated_pairs_with_results.csv") %>%
  mutate(
    # Recode pronoun as 'extrashort_definite' before separating
    recipient_id = case_when(
      recipient_id == 'shortDefinite' ~ 'short_definite',
      recipient_id == 'longDefinite' ~ 'long_definite',
      recipient_id == 'shortIndefinite' ~ 'short_indefinite',
      recipient_id == 'longIndefinite' ~ 'long_indefinite',
      recipient_id == 'pronoun' ~ 'extrashort_definite'
    )
  ) %>%
  separate(recipient_id, 
           into = c('recipient_length', 'recipient_definiteness'), 
           sep = '_') %>%
  mutate(
    # Set reference levels (optional - R will use alphabetical by default)
    recipient_length = factor(recipient_length, 
                             levels = c("extrashort", "short", "long")),
    recipient_definiteness = factor(recipient_definiteness,
                                   levels = c("definite", "indefinite"))
  )

df_merged = df_judgments %>%
  mutate(do_sent = DOsentence,
         pd_sent = PDsentence) %>%
  inner_join(df_pythia_models %>% filter(step == 143000)) 

coef_results = df_merged %>%
  group_by(mpath, n_params, seed_name) %>%
  group_modify(~broom::tidy(lm(log_odds_total ~ classification + recipient_length +
                                 recipient_definiteness, data = .x))) %>%
  filter(term != "(Intercept)")


coef_summary <- coef_results %>%
  group_by(mpath, n_params, term) %>%
  summarise(
    mean_est = mean(estimate),
    sd_est = sd(estimate),
    se_est = sd(estimate) / sqrt(n())
  )

library(ggstance)

ggplot(coef_results, aes(y = term, x = estimate)) +
  geom_pointrange(data = coef_summary,
                  aes(x = mean_est, 
                      xmin = mean_est - 1.96*se_est, 
                      xmax = mean_est + 1.96*se_est),
                  size = 0.8, linewidth = 1.2) +
  geom_point(aes(color = seed_name), 
             position = position_jitter(height = 0.2),
             size = 2, alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Coefficient estimate",
    y = "Model / Term",
    color = "Seed"
  ) +
  theme_minimal() +
  facet_wrap(~mpath)



ggplot(coef_results, aes(x = seed_name, y = estimate, group = term, color = term)) +
  geom_line(size = 1) +
  geom_point(size = 3) +
  facet_wrap(~mpath, scales = "free_x") +
  labs(
    x = "Seed",
    y = "Coefficient estimate",
    color = "Term"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




mod = lmer(data = df_merged %>%
             filter(recipient_id %in% c("longIndefinite", "shortIndefinite")),
           log_odds_mean ~ recipient_id + (1 + classification|seed_name))
summary(mod)


mod = lmer(data = df_merged %>%
             filter(recipient_id %in% c("longDefinite", "shortDefinite")),
           log_odds_mean ~ recipient_id + (1 + classification|seed_name))
summary(mod)


df_merged %>%
  group_by(mpath, seed_name, recipient_length) %>%
  summarise(mean_lr_total = mean(log_odds_total),
            se_lr_total = sd(log_odds_total)/sqrt(n())) %>%
  ggplot(aes(x = reorder(recipient_length, mean_lr_total),
             y = mean_lr_total)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(x = reorder(recipient_length, mean_lr_total),
                    ymin = mean_lr_total - se_lr_total,
                    ymax = mean_lr_total + se_lr_total),
                width = .2) +
  theme_minimal() +
  facet_wrap(~mpath + seed_name)

# Calculate mean across seeds for each model/length/definiteness combo
df_seed_means <- df_merged %>%
  group_by(mpath, n_params, recipient_length, recipient_definiteness) %>%
  summarise(
    mean_lr_total = mean(log_odds_total),
    se_lr_total = sd(log_odds_total) / sqrt(n()),
    .groups = "drop"
  )

# Calculate individual seed values
df_seed_values <- df_merged %>%
  group_by(mpath, n_params, seed_name, recipient_length, recipient_definiteness) %>%
  summarise(mean_lr_total = mean(log_odds_total), .groups = "drop")


ggplot() +
  # Individual seeds - smaller points with jitter
  geom_point(data = df_seed_values,
             aes(x = mean_lr_total, 
                 y = reorder(recipient_length, mean_lr_total),
                 color = recipient_definiteness),
             position = position_jitter(height = 0.15, seed = 42),
             size = 2, alpha = 0.5) +
  # Mean across seeds - larger points with error bars
  geom_pointrange(data = df_seed_means,
                  aes(x = mean_lr_total,
                      xmin = mean_lr_total - se_lr_total,
                      xmax = mean_lr_total + se_lr_total,
                      y = reorder(recipient_length, mean_lr_total),
                      color = recipient_definiteness),
                  size = 1, linewidth = 1.2,
                  position = position_dodge(width = 0.3)) +  # Changed to width
  labs(
    x = "Log odds (DO preference)",
    y = "Recipient length",
    color = "Recipient\nDefiniteness"
  ) +
  facet_wrap(~mpath, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "right")



#### Human baseline

mod = lmer(data = df_judgments,
           BehavDOpreference ~ classification + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(recipient_id %in% c("longIndefinite", "shortIndefinite")),
           BehavDOpreference ~ recipient_id + (1 + recipient_id|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(recipient_id %in% c("longDefinite", "shortDefinite")),
           BehavDOpreference ~ recipient_id + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments %>%
             filter(theme_type != "something"),
           BehavDOpreference ~ theme_type + (1 + classification|verb_id))
summary(mod)

mod = lmer(data = df_judgments,
           BehavDOpreference ~ classification + (1 + classification|verb_id))
summary(mod)


df_pythia_models %>%
  mutate(lr = pd_surprisal_total - do_surprisal_total) %>%
  ggplot(aes(x = lr,
             log_odds_total)) +
  geom_point()
```


